---
title: "Survival analyses"
author: "Vanessa Dumeaux"
date: "2024-04-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load functions to conduct survival analyses


```{r}
source("../../src/survival.R")
```


# Load packages
```{r}
library(reshape2)
library(survival)
library(survminer)
library(maftools)
library(coxphf)
library(clusterProfiler)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(enrichplot)
```


```{r}

# Load data
dcis <- readRDS("/path/to/Rdata/output/exp04/dcis.rds")

# Survival analysis all patients
all.sig.genes <- survival.analysis(maf_data = dcis, time = "time.10", event = "event.10", clinsubset = NULL, top = nrow(dcis@gene.summary[dcis@gene.summary$MutatedSamples>5]), pvalue = 0.01)

# Survival analysis RT treated patients
radiation.sig.genes <- survival.analysis(maf_data = dcis, time = "time.10", event = "event.10", clinsubset = "radiation==TRUE", top = nrow(dcis@gene.summary[dcis@gene.summary$MutatedSamples>5]), pvalue = 0.05)

# Survival analysis non-RT treated patients
noradiation.sig.genes <- survival.analysis(maf_data = dcis, time = "time.10", event = "event.10", clinsubset = "radiation==FALSE", top = nrow(dcis@gene.summary[dcis@gene.summary$MutatedSamples>5]), pvalue = 0.05)

```


```{r}

# Variant classification colours
vc_cols = RColorBrewer::brewer.pal(n = 8, name = 'Paired')
names(vc_cols) = c(
  'Frame_Shift_Del',
  'Missense_Mutation',
  'Nonsense_Mutation',
  'Multi_Hit',
  'Frame_Shift_Ins',
  'In_Frame_Ins',
  'Splice_Site',
  'In_Frame_Del'
)

# Annotation colours
col.clinical <- list()
col.clinical[["subtype"]] <- c(LumA="blue4", LumB="deepskyblue", Basal="firebrick2", Normal="green4", Her2="hotpink2", "NA"="darkgrey")
col.clinical[["age"]] <- c(over60="#3182bd", "50-60"="#9ecae1", under50="#deebf7")
col.clinical[["nugradef"]] <- c(High="#de2d26", Moderate="#fc9272", Low="#fee0d2")
col.clinical[["event.10.chr"]] <- c("0"="#fee0b6", "1"="#b35806")


# Oncoplot all significant genes
pdf("../../figures/exp07/oncoplot_all_genes_coxphf.pdf", width =12)
oncoplot(maf = all.sig.genes$maf, genes = all.sig.genes$genes, colors = vc_cols, clinicalFeatures = c('event.10.chr', 'age', 'nugradef','subtype'), annotationColor=col.clinical,fontSize = 0.3,  gene_mar = 8, showTitle = FALSE)
dev.off()

# Oncoplot for RT associated genes
pdf("../../figures/exp07/oncoplot_radiation_genes_coxphf.pdf", width =12)
oncoplot(maf = radiation.sig.genes$maf, genes = radiation.sig.genes$genes, colors = vc_cols, clinicalFeatures = c('event.10.chr','age', 'nugradef','subtype'), annotationColor=col.clinical,fontSize = 0.3,  gene_mar = 8, showTitle = FALSE)
dev.off()




```
# More exploratory plots for genes significantly associated with LR risk in all patients 

## HR plots ordered like oncoplot

```{r}

# filter significant genes for plotting
plotdf <- all.sig.genes$surv_result$restb[all.sig.genes$surv_result$restb$Gene_combination %in% all.sig.genes$genes,]

# Set factor levels to have the same order than oncoplot
plotdf$Gene_combination <- factor(plotdf$Gene_combination, 
                                     levels = rev(c("SH2B2", "MYO7A", "MUCL3", "DNASE2B", "PDZD8")), ordered = TRUE)
                                     
# Hazard ratio plot                                     
pdf(paste0("../../figures/exp07/hr_allgenes_ordered.pdf"), height = 10, width = 20)
ggplot(plotdf, aes(y = Gene_combination)) +
         geom_point(aes(x = hr), size = 3) +
         geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
         geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
         scale_x_log10() +
         theme_bw() +
         theme(
           panel.grid.minor = element_blank(),
           panel.grid.major.y = element_blank()
         ) +
         labs(
           x = "",
           y = "",
           title = ""
         ) +
         # Add HR and CI values as text
         geom_text(aes(x = max(conf.high) * 1.2, 
                       label = sprintf("%.2f (%.2f-%.2f)", hr, conf.low, conf.high)),
                   hjust = 0) +
         # Add p-values
         geom_text(aes(x = max(conf.high) * 2, 
                       label = sprintf("p = %.3f", P_value)),
                   hjust = 0)
dev.off()
```

## KM curves
```{r}
# Survival analysis on significant genes (gene set size =  12)
combo.all.sig.genes <- survival.analysis(maf_data = all.sig.genes$maf, 
                                         genes = all.sig.genes$genes, 
                                         geneSetSize = 5, 
                                         genecombo = 'union', minMutations=1,
                                         time = "time.10", 
                                         event = "event.10", pvalue = 0.01) 

# Kaplan-Meier plot: No patients in the radiated group without at least one mutation in these 13 genes experienced recurrence within 10 years
km <- ggsurvplot(
      combo.all.sig.genes$surv_result$res$PDZD8_SH2B2_MYO7A_DNASE2B_MUCL3$fit.km, 
      # survfit object with calculated statistics.
      data=combo.all.sig.genes$surv_result$res$MPDZD8_SH2B2_MYO7A_DNASE2B_MUCL3$data,
      pval = TRUE,             # show p-value of log-rank test.
      conf.int = TRUE,         # show confidence intervals for 
                            # point estimaes of survival curves.
      conf.int.style = "step",  # customize style of confidence intervals
      xlab = "Time in years",   # customize X axis label.
      break.time.by = 3,     # break X axis in time intervals by 200.
      ggtheme = theme_light(), # customize plot and risk table with a theme.
      risk.table = "abs_pct",  # absolute number and percentage at risk.
      risk.table.y.text.col = T,# colour risk table text annotations.
      risk.table.y.text = FALSE,# show bars instead of names in text annotations
                            # in legend of risk table.
      ncensor.plot = TRUE,      # plot the number of censored subjects at time t
      surv.median.line = "hv",  # add the median survival pointer.
      title = ""
   )
km

pdf("../../figures/exp07/km_union_all_genes_coxphf.pdf")
print(km)
dev.off()
```




## GO annotations - significant genes in all patients

```{r}

# Load gene ontology data for molecular function (MF)
hsGO <- godata('org.Hs.eg.db', ont="MF")

# Convert gene symbols to Entrez IDs
entrez_ids <- mapIds(org.Hs.eg.db,
                    keys = as.character(plotdf$Gene_combination),
                    column = "ENTREZID",
                    keytype = "SYMBOL",
                    multiVals = "first")

# Perform gene ontology classification for cellular component (CC)
ggo <- groupGO(gene     = entrez_ids,
               OrgDb    = org.Hs.eg.db,
               ont      = "CC",
               level    = 3,
               readable = TRUE)

# Convert GO results to dataframe and filter for terms with at least one associated gene  
ggodf <- as.data.frame(ggo)
ggodf[ggodf$Count>0,]

# Expand gene lists within GO terms and mark presence  
genes_df <- ggodf[ggodf$Count>0,] %>%
  separate_rows(geneID, sep = "/") %>%
  mutate(Present = 1)

# GO enrichment plot for Cellular Component terms 
allgenes.cc.plot <- ggplot(genes_df, aes(x = Description, y = geneID)) +
  geom_point(size = 3, color = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(x = "Cellular Component GO Terms",
       y = "",
       title = "") +
  theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
  axis.text.y = element_text(size = 10)
)

ggsave("../../figures/exp07/allgenes_cc_plot.pdf", allgenes.cc.plot)

```


# More exploratory plots for genes significantly associated with LR risk in RT-treated patients 


## KM curves - combined variant radiation


```{r}

# Survival analysis on radiation-associated significant genes
combo.radiation.sig.genes <- survival.analysis(maf_data =radiation.sig.genes$maf, 
                                               genes = radiation.sig.genes$genes, 
                                               geneSetSize = 29, genecombo = 'union',
                                               minMutations = 2,
                                               time = "time.10", 
                                               event = "event.10", 
                                               pvalue = 0.05) 
# Note: no event within 10 years for radiated group which do not have at least one mutation in one of these 13 genes

# Generate Kaplan-Meier survival plot for the identified gene combination
km <- ggsurvplot(
      combo.radiation.sig.genes$surv_result$res$ANO6_SH2B2_REV1_C6orf118_MUCL3_PCSK1_WWC1_TMEM132C_SORBS1_CTBP2_TTLL10_KRT1_EML5_MAGEC3_EXD3_NPFFR1_ABCA2_MGAM2_CEP128_GLI3_PTPN13_CFAP47_TBC1D4_ZNF641_AADACL3_AFF3_MROH2B_CALU_KIF24$fit.km,
      data=combo.radiation.sig.genes$surv_result$res$ANO6_SH2B2_REV1_C6orf118_MUCL3_PCSK1_WWC1_TMEM132C_SORBS1_CTBP2_TTLL10_KRT1_EML5_MAGEC3_EXD3_NPFFR1_ABCA2_MGAM2_CEP128_GLI3_PTPN13_CFAP47_TBC1D4_ZNF641_AADACL3_AFF3_MROH2B_CALU_KIF24$data,
      pval = TRUE,             # show p-value of log-rank test.
      conf.int = TRUE,         # show confidence intervals for 
                            # point estimaes of survival curves.
      conf.int.style = "step",  # customize style of confidence intervals
      xlab = "Time in years",   # customize X axis label.
      break.time.by = 3,     # break X axis in time intervals by 200.
      ggtheme = theme_light(), # customize plot and risk table with a theme.
      risk.table = "abs_pct",  # absolute number and percentage at risk.
      risk.table.y.text.col = T,# colour risk table text annotations.
      risk.table.y.text = FALSE,# show bars instead of names in text annotations
                            # in legend of risk table.
      ncensor.plot = TRUE,      # plot the number of censored subjects at time t
      surv.median.line = "hv",  # add the median survival pointer.
      title = ""
   )

pdf("../../figures/exp07/km_union_radiation_genes_coxphf.pdf")
print(km)
dev.off()
```


## Survival analysis with Cox proportional hazards model, adjusting for age

```{r}

# Check cox (also after adjustment for age - use Firth's penalized likelihood Cox regression)
cd <- combo.radiation.sig.genes$surv_result$res$ANO6_SH2B2_REV1_C6orf118_MUCL3_PCSK1_WWC1_TMEM132C_SORBS1_CTBP2_TTLL10_KRT1_EML5_MAGEC3_EXD3_NPFFR1_ABCA2_MGAM2_CEP128_GLI3_PTPN13_CFAP47_TBC1D4_ZNF641_AADACL3_AFF3_MROH2B_CALU_KIF24$data
firth_model <- coxphf(Surv(Time, Status) ~ Group+age, data = cd)
firth_model

# Perform survival analysis on radiation-specific genes but when including all patients
combo.radiation.sig.genes.forall <- survival.analysis(
  maf_data = all.sig.genes$maf,
  genes = radiation.sig.genes$genes[!radiation.sig.genes$genes %in% intersect(radiation.sig.genes$genes, all.sig.genes$genes)], 
  geneSetSize = 27, genecombo = 'union', minMutations =2, 
  time = "time.10", event = "event.10", pvalue = 0.01) # not significant when all patients are included

```

## check if it's related with number of synonymous variable

```{r}

# Extract survival data
dat <- combo.radiation.sig.genes$surv_result$res$ANO6_REV1_C6orf118_PCSK1_WWC1_TMEM132C_SORBS1_CTBP2_TTLL10_KRT1_EML5_MAGEC3_EXD3_NPFFR1_ABCA2_MGAM2_CEP128_GLI3_PTPN13_CFAP47_TBC1D4_ZNF641_AADACL3_AFF3_MROH2B_CALU_KIF24$data
dat$n_nonsynonymous_mut <- as.numeric(dat$n_nonsynonymous_mut)

# T-test to compare non synonymous mutation burden between groups
t.test(dat$n_nonsynonymous_mut~dat$Group)

# Violin plot of non synonymous mutation burden across groups
pdf("../../figures/exp07/union_radiation_genes_nMut_violin_coxphf.pdf", height = 4)
ggplot(dat, aes(x = Group, y = log(n_nonsynonymous_mut))) +
  geom_violin(alpha = 0.5) +
  geom_point(aes(x = Group, y = log(n_nonsynonymous_mut), colour = event.10.chr),position = position_jitter(seed = 1, width = 0.2)) +
  scale_colour_manual(values = c("#fee0b6", "#b35806")) +
  theme(legend.position = "none") +
  theme_minimal() +
  labs(x="", y="# non-synonymous variants (log2)", color="local recurrence")
dev.off()

```


## GO annotations. - significant genes in radiation-treated patients
```{r}
plotdf <- radiation.sig.genes$surv_result$restb[radiation.sig.genes$surv_result$restb$Gene_combination %in% radiation.sig.genes$genes,]


# Convert gene symbols to Entrez IDs
entrez_ids <- mapIds(org.Hs.eg.db,
                    keys = as.character(plotdf$Gene_combination),
                    column = "ENTREZID",
                    keytype = "SYMBOL",
                    multiVals = "first")

# GO enrichment for Biological Process (BP)  
ego10 <- enrichGO(entrez_ids, 
                OrgDb = 'org.Hs.eg.db', ont="BP", pvalueCutoff=1, readable = TRUE)

# WikiPathways (WP) enrichment analysis  
ego10 <- enrichWP(entrez_ids, pvalueCutoff=1,  organism = "Homo sapiens")
ego10[ego10$qvalue < 0.2,] 

# GO enrichment analysis for Cellular Component (CC)  
ego10 <- enrichGO(entrez_ids,
                OrgDb = 'org.Hs.eg.db', 
                ont="CC", pvalueCutoff = 1, 
                readable = TRUE, 
                qvalueCutoff = 0.2)

# Compute pairwise similarity between enriched GO terms
ego <- pairwise_termsim(ego10)

# Enrichment map plot for Cellular Component terms 
pdf("../../figures/exp07/radiationgenes_cc_net.pdf", width = 10)
emapplot(ego, showCategory = 100)  # adjust number of categories as needed
dev.off()


```