---
title: "CNApp"
author: "Vanessa Dumeaux"
date: "2024-09-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# CNapp to conduct region analyses

We used CNapp to apply a re-segmentation approach to adjust for amplitude divergence due to technical variability using the default settings (minimum segment length = 100 Kbp, minimum amplitude deviation from segment to zero = 0.16, maximum distance between segments=1 Mb, maximum amplitude deviation between segments = 0.16, and maximum BAF deviation between segments = 0.1). 


# Reformat ASCAT output for CNApp

```{r}
ascat_output <- readRDS("/path/to/Rdata/output/exp06/output_ascat/Robj/ascat.output.rds")
ascat_ploidy <- read.table("/path/to/Rdata/output/exp06/output_ascat/ascat_ploidy_filtered.tsv", sep="\t", header = TRUE)


# Merge CN data with ploidy data
cn <- ascat_output %>%
  left_join(ascat_ploidy, by = "sample") %>%
  mutate(relative_cn = (ascat_output$nMajor + ascat_output$nMinor)/ploidy) %>%
  mutate(absolute_cn = ascat_output$nMajor + ascat_output$nMinor)
  
cn$absolute_cn_modified <- ifelse(cn$absolute_cn >= 5, 5, cn$absolute_cn)   


cnapp <- data.frame(
  ID = cn$sample,
  chr = cn$chr,
  loc.start = cn$startpos,
  loc.end = cn$endpos,
  seg.mean = log2(cn$absolute_cn) - log2(cn$ploidy))

write.table(cnapp, "/path/to/Rdata/output/exp06/cnapp_input.txt", sep="\t", quote=FALSE, row.names=FALSE)

dcis <- readRDS("/path/to/Rdata/output/exp04/dcis.rds")

library(tibble)
sample_annotation <- dcis@clinical.data
sample_annotation$ID <- sample_annotation$Tumor_Sample_Barcode

df <- sample_annotation %>%
  select(ID, everything())

df$surv_status <- df$event.10
df$surv_time <- df$time.10


write.table(df, "/path/to/Rdata/output/exp06/cnapp_clinical_input.txt", sep="\t", quote=FALSE, row.names=FALSE)


```



## Run CNapp Shiny application

```{r}

if(!require(shiny)) install.packages("shiny")
if(!require(shinyBS)) install.packages("shinyBS")
if(!require(shinyjs)) install.packages("shinyjs")
if(!require(shinythemes)) install.packages("shinythemes")
if(!require(shinyWidgets)) install.packages("shinyWidgets")
if(!require(shinydashboard)) install.packages("shinydashboard")
if(!require(survival)) install.packages("survival")
if(!require(survminer)) install.packages("survminer")
if(!require(Cairo)) install.packages("Cairo")


if(!require(V8)) install.packages("V8")
#if any issue here like 'ERROR: configuration failed for package ‘curl’' or 'ERROR: configuration failed for package ‘V8’' follow printed instructions on your command-line window

if(!require(httr)) install.packages("httr")
#if any issue here like 'ERROR: configuration failed for package ‘openssl’' follow printed instructions on your command-line window

if(!require(plotly)) install.packages("plotly")
if(!require(randomcoloR)) install.packages("randomcoloR")
if(!require(heatmaply)) install.packages("heatmaply")
if(!require(ggplot2)) install.packages("ggplot2")
if(!require(ggsignif)) install.packages("ggsignif")
if(!require(RColorBrewer)) install.packages("RColorBrewer")
if(!require(randomForest)) install.packages("randomForest")
if(!require(doBy)) install.packages("doBy")
if(!require(parallel)) install.packages("parallel")
if(!require(caret)) install.packages("caret")
if(!require(XML)) install.packages("XML")
#if any issue here like 'ERROR: configuration failed for package ‘XML’' ty installing 'libxml2' into your OS

if (!require(devtools)) install.packages("devtools")

if(!require(limma)) {source("http://www.bioconductor.org/biocLite.R");biocLite("limma")}
if(!require(GenomicFeatures)) {source("http://www.bioconductor.org/biocLite.R");biocLite("GenomicFeatures")}
#if any issue here like 'ERROR: configuration failed for package ‘RMySQL’' ty installing 'libmysqlclient' into your OS
if(!require(GenomicAlignments)) {source("http://www.bioconductor.org/biocLite.R");biocLite("GenomicAlignments")}
if(!require(GenVisR)) {source("http://www.bioconductor.org/biocLite.R");biocLite("GenVisR")}

    ############## to use BiocManager ####################
    #if (!requireNamespace("BiocManager", quietly = TRUE))
    #    install.packages("BiocManager")
    #BiocManager::install("limma")
    #BiocManager::install("GenomicFeatures")
    #BiocManager::install("GenomicAlignments")
    #BiocManager::install("GenVisR")
    ######################################################

if(!require(shinysky)) {library(devtools); devtools::install_github("AnalytixWare/ShinySky")}
#if(!require(GenVisR)) {library(devtools); devtools::install_github("griffithlab/GenVisR")}

if (!require(webshot)) install.packages("webshot")

# if(!require(phantomjs)) {webshot::install_phantomjs()}

```



```{r}
# Easiest way is to use runGitHub to run CNApp from GitHub:
library(shiny)
runGitHub("CNApp", "ait5")
```

1. Re-segmentation approach to adjust for amplitude divergence due to technical variability using the default settings (minimum segment length = 100 Kbp, minimum amplitude deviation from segment to zero = 0.16, maximum distance between segments=1 Mb, maximum amplitude deviation between segments = 0.16, and maximum BAF deviation between segments = 0.1)
2. Run Region Analyses for cytobands and save outputs

# GCS analyses
```{r}
cytoband <- read.table("/path/to/Rdata/output/exp06/cnapp/region_analyses/cytobands/cna_profile_Cytobands_by_ID2024-09-12 10_47_01.867477.tsv", sep="\t", header = TRUE, row.names = 1)

cl <- read.table("/path/to/Rdata/output/exp06/cnapp/res-seg_cna-scores/allsamples_CNApp_CNA_Scores_&_annotations_2024-09-10 10_45_15.475147.tsv", sep="\t", header = TRUE, row.names = 1)

cl.df <- cl[ , c('event.10.chr','age', 'nugradef','subtype', 'GCS', 'time.10')]

col.clinical <- list()
col.clinical[["subtype"]] <- c(LumA="blue4", LumB="deepskyblue", Basal="firebrick2", Normal="green4", Her2="hotpink2", "NA"="darkgrey")
col.clinical[["age"]] <- c(over60="#3182bd", "50-60"="#9ecae1", under50="#deebf7")
col.clinical[["nugradef"]] <- c(High="#de2d26", Moderate="#fc9272", Low="#fee0d2")
col.clinical[["event.10.chr"]] <- c("0"="#fee0b6", "1"="#b35806")
col.clinical[["GCS"]] <- viridisLite::viridis(n=100, option="G", direction = -1)

table(colnames(cytoband)==rownames(cl.df))
cytoband <- cytoband[, rownames(cl.df)[order(cl.df$GCS)]]


# Set custom color breaks for relative copy number values with unique breaks
color_breaks <- unique(c(seq(-3, -0.5, length.out = 25), 
                         seq(-0.5, 0, length.out = 25), 
                         seq(0, 0.5, length.out = 25), 
                         seq(0.5, 3, length.out = 25)))
# Generate heatmap with clustering and annotations
pheatmap::pheatmap(cytoband, cluster_cols = FALSE, cluster_rows = FALSE, annotation_col = cl.df,
         color = colorRampPalette(c("blue", "white", "red"))(length(color_breaks) - 1),
         breaks = color_breaks,
         show_colnames = FALSE,  show_rownames = TRUE,
         main = "Relative Copy Number Alterations Heatmap",
         annotation_colors = col.clinical, filename = "../../figures/exp06/rel_cn_cnapp_cytoband_heatmap.pdf")
```

## GCS across grade
```{r}
library(ggridges)
library(ggplot2)
library(ggpubr)
library(viridis)
library(hrbrthemes)


# Grade
# Reorder the factor levels
cl.df$nugradef <- factor(cl.df$nugradef, levels = c("Low", "Moderate", "High"))

p6c <- ggplot(cl.df, aes(x = nugradef, y = GCS, fill = nugradef)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white", alpha = 0.5) +
  geom_jitter(width = 0.1, alpha = 0.5, size = 1) +
  scale_fill_manual(values = c("Low" = "#fee0d2", 
                               "Moderate" = "#fc9272", 
                               "High" = "#de2d26")) +
  labs(x = "Grade",
       y = "GCS Score") +
  theme_minimal() +
  theme(legend.position = "none") +
  stat_compare_means(comparisons = list(c("Low", "High")),
                     label = "p.signif",
                     method = "t.test",
                     label.y = max(cl.df$GCS) + 3)


t.test(cl.df$GCS[cl.df$nugradef == "High"], 
                          cl.df$GCS[cl.df$nugradef == "Moderate"])
t.test(cl.df$GCS[cl.df$nugradef == "High"], 
                     cl.df$GCS[cl.df$nugradef == "Low"])
t.test(cl.df$GCS[cl.df$nugradef == "Moderate"], 
                         cl.df$GCS[cl.df$nugradef == "Low"])

ggsave("../../figures/exp06/grade_gcs_violin.pdf", p6c, height = 5)
```





## GCS across subtypes
```{r}

# Define the color scheme
subtype_colors <- c(LumA="blue4", LumB="deepskyblue", Basal="firebrick2", 
                    Normal="green4", Her2="hotpink2")

# Remove NA subtype and reorder factor levels
cl.df_filtered <- cl.df[!is.na(cl.df$subtype), ]
cl.df_filtered$subtype <- factor(cl.df_filtered$subtype, 
                                 levels = c("Normal", "LumA", "LumB", "Her2", "Basal"))

# Perform pairwise t-tests
subtype_pairs <- combn(levels(cl.df_filtered$subtype), 2, simplify = FALSE)
p_values <- sapply(subtype_pairs, function(pair) {
  t.test(cl.df_filtered$GCS[cl.df_filtered$subtype == pair[1]],
         cl.df_filtered$GCS[cl.df_filtered$subtype == pair[2]])$p.value
})

# Filter significant comparisons
sig_comparisons <- subtype_pairs[p_values < 0.05]

# Calculate y positions for significance bars
max_y <- max(cl.df_filtered$GCS)
y_positions <- seq((max_y) * 1.55, (max_y) * 1.75, length.out = length(sig_comparisons))

# Create the plot
p6d <- ggplot(cl.df_filtered, aes(x = subtype, y = GCS, fill = subtype)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white", alpha = 0.5) +
  geom_jitter(width = 0.1, alpha = 0.5, size = 1) +
  scale_fill_manual(values = subtype_colors) +
  labs(x = "Subtype",
       y = "GCS Score") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  stat_compare_means(comparisons = sig_comparisons,
                     label = "p.signif",
                     method = "t.test",
                     label.y = y_positions, tip.length = 0.01,
                     vjust = 0.5)

p6d

ggsave("../../exp06/subtype_gcs_violin.pdf", p6d, height = 5)
```



# Arms gain/loss enrichment significance analysis ----------
```{r}

arms <- read.table("/data/lab_vm/refined/bc_dcis/DNAseq_allBatches/R/exp39/cnapp/region_analyses/chrom_arms/cna_profile_Arms_by_GCS2024-09-12 10_06_19.168891.tsv", sep="\t", header = TRUE, row.names = 1)


arms_long <- rownames_to_column(arms, var = "arms")
# Convert from wide to long format
arms_long <- pivot_longer(
  arms_long,
  cols = starts_with("DCRT"),      # Columns to pivot
  names_to = "ID",                          # Name of the new key column
  values_to = "cna"          # Name of the new value column
)

cl.df2 <- rownames_to_column(cl.df, var = "ID")

arms_long <- left_join(arms_long, cl.df2)
```
## plot proportions of gains/losses across arms incl all samples
```{r}

# Load necessary packages
library(dplyr)
library(tidyr)
library(ggplot2)

# Step 1: Extract chromosome number and arm
arms_long <- arms_long %>%
  mutate(
    chr_num = gsub("chr(\\d+|X|Y)_.*", "\\1", arms),
    chr_arm = gsub(".*_(p|q)", "\\1", arms)
  )

# Step 2: Convert chromosome numbers to numeric, handling X and Y chromosomes
arms_long <- arms_long %>%
  mutate(
    chr_num = case_when(
      chr_num == "X" ~ 23,
      chr_num == "Y" ~ 24,
      TRUE ~ as.numeric(chr_num)
    )
  )
```


## Get stats to see if arms' gain and losses are significantly more frequent to what is expected by chance (taking overall gain/losses  proportion across all arms)

```{r}
# Calculate total number of samples
total_samples <- n_distinct(arms_long$ID)

# Calculate gains and losses proportions for each arm
arm_summary <- arms_long %>%
  group_by(arms) %>%
  summarise(
    gains = sum(cna > 0.2, na.rm = TRUE),
    losses = sum(cna < -0.2, na.rm = TRUE),
    total = n_distinct(ID)
  ) %>%
  mutate(
    gain_prop = gains / total,
    loss_prop = losses / total
  )

# Rank arms by gain proportion
most_frequent_gains <- arm_summary %>%
  arrange(desc(gain_prop))

# Rank arms by loss proportion
most_frequent_losses <- arm_summary %>%
  arrange(desc(loss_prop))

# View top 10 most frequent gains and losses
top_gains <- head(most_frequent_gains, 10)
top_losses <- head(most_frequent_losses, 10)

# Calculate overall gain and loss proportions
overall_gain_prop <- sum(arm_summary$gains) / (total_samples * nrow(arm_summary))
overall_loss_prop <- sum(arm_summary$losses) / (total_samples * nrow(arm_summary))

# Perform binomial tests for gains
arm_summary <- arm_summary %>%
  rowwise() %>%
  mutate(
    gain_p_value = binom.test(
      x = gains,
      n = total,
      p = overall_gain_prop,
      alternative = "greater"
    )$p.value,
    loss_p_value = binom.test(
      x = losses,
      n = total,
      p = overall_loss_prop,
      alternative = "greater"
    )$p.value
  )

# Adjust p-values for multiple testing
arm_summary <- arm_summary %>%
  mutate(
    gain_p_adj = p.adjust(gain_p_value, method = "BH"),
    loss_p_adj = p.adjust(loss_p_value, method = "BH")
  )

# Identify significant gains and losses
significant_gains <- arm_summary %>%
  filter(gain_p_adj < 0.01) %>%
  arrange(gain_p_adj)

significant_gains

significant_losses <- arm_summary %>%
  filter(loss_p_adj < 0.01) %>%
  arrange(loss_p_adj)

significant_losses

```




## compute the same statistics but per subtype

Investigate whether gain/loss in arms are specific for one subtype compared to other subtypes (fisher exact tests)

```{r}
# Load necessary packages
library(dplyr)


# Corrected function to perform Fisher's exact test for a given arm and subtype
fisher_test_arm_subtype <- function(data, arm, subtype_of_interest, alteration_type = c("Gain", "Loss"), threshold = 0.2) {
  alteration_type <- match.arg(alteration_type)
  
  # Subset data for the specific arm
  data_arm <- data[data$arms == arm,]
  
  # Define alteration condition based on type, within data_arm
  if (alteration_type == "Gain") {
    alteration_condition <- data_arm$cna > threshold
  } else {
    alteration_condition <- data_arm$cna < -threshold
  }

  # Create variables indicating presence of alteration and subtype membership
  data_arm <- data_arm %>%
    mutate(
      alteration = ifelse(alteration_condition, "Altered", "Not_Altered"),
      subtype_group = ifelse(subtype == subtype_of_interest, subtype_of_interest, "Other")
    )
  
  # Create the contingency table
  contingency_table <- table(data_arm$alteration, data_arm$subtype_group)
  
  # Check if the table has the correct dimensions
  if (all(dim(contingency_table) == c(2, 2))) {
    # Perform Fisher's exact test
    fisher_result <- fisher.test(contingency_table)
    
    # Return the results
    return(data.frame(
      arm = arm,
      subtype = subtype_of_interest,
      alteration_type = alteration_type,
      p_value = fisher_result$p.value,
      odds_ratio = fisher_result$estimate
    ))
  } else {
    # Return NA if the contingency table doesn't have both categories
    return(data.frame(
      arm = arm,
      subtype = subtype_of_interest,
      alteration_type = alteration_type,
      p_value = NA,
      odds_ratio = NA
    ))
  }
}

# Get the list of unique arms and subtypes
arms_list <- unique(arms_long$arms)
subtypes_list <- unique(arms_long$subtype)

# Remove NA subtypes if present
subtypes_list <- subtypes_list[!is.na(subtypes_list)]

# Initialize an empty data frame to store results
fisher_results <- data.frame()

# Loop over each alteration type
for (alteration in c("Gain", "Loss")) {
  # Loop over each arm
  for (arm in arms_list) {
    # Loop over each subtype
    for (subtype in subtypes_list) {
      # Perform Fisher's exact test for the current arm, subtype, and alteration
      result <- fisher_test_arm_subtype(
        data = arms_long,
        arm = arm,
        subtype = subtype,
        alteration_type = alteration,
        threshold = 0.2  # Adjust threshold if necessary
      )
      # Append the result to the results data frame
      fisher_results <- rbind(fisher_results, result)
    }
  }
}


# Adjust p-values using the Benjamini-Hochberg method
fisher_results <- fisher_results %>%
  mutate(
    p_adj = p.adjust(p_value, method = "BH")
  )

# View significant results (e.g., p_adj < 0.05)
significant_results <- fisher_results %>%
  filter(!is.na(p_value) & p_value < 0.01) %>%
  arrange(p_value)

# View the significant results
significant_results


```



